A unified pseudo-$C_\ell$ framework

### 2018-08-16 doc/Paper/main.tex
Addressed some of Boris' and Scott's comments

% Add your own macros here:
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{dsfont}

\lstset{ %
language=python,                % choose the language of the code
basicstyle=\footnotesize,       % the size of the fonts that are used for the code
numbers=left,                   % where to put the line-numbers
numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
frame=false,           % adds a frame around the code
tabsize=2,          % sets default tabsize to 2 spaces
captionpos=b,           % sets the caption-position to bottom
breaklines=true,        % sets automatic line breaking
breakatwhitespace=false,    % sets if automatic breaks should only happen at whitespace
escapeinside={\%*}{*)}          % if you want to add a comment within your code
}
%\usepackage[colorlinks,linkcolor=red,citecolor=blue,urlcolor=blue ]{hyperref}
\newcommand{\nv}{\hat{\bf n}}
\newcommand{\wtj}[6]{\left(\begin{array}{ccc} #1 & #2 & #3\\#4 & #5 & #6\end{array} \right)}
\newcommand{\nmt}{{\tt NaMaster}}
\newcommand{\va}{{\mathbf{a}}}
\newcommand{\vb}{{\mathbf{b}}}
\newcommand{\mC}{{\sf C}}
\newcommand{\mM}{{\sf M}}
\newcommand{\mI}{{\mathds{1}}}

@@

\begin{abstract}
  The pseudo-$C_\ell$ is an algorithm for estimation of the angular
  power and cross-power spectra that is very fast and in realistic
  cases also nearly optimal. The algorithm can be extended to deal
  with contaminant deprojection and $E/B$ purification, and can
  therefore be applied in a wide variety of scenarios of interest for
  current and future cosmological observations. This paper presents
  \nmt, a public, validated, robust, accurate and easy-to-use software
  package that, for the first time, provides a unified framework to
  compute angular cross-power spectra of any pair of spin-0 or spin-2
  fields, contaminated by an arbitrary number of linear systematics
  and requiring $B$- or $E$-mode purification, both in the sphere or
  in the flat-sky approximation. We describe the mathematical
  background of the estimator, including all the features above, and
  its software implementation in \nmt. We construct a validation suite
  that aims to resemble the types of observations that next-generation
  large-scale structure and ground-based CMB experiments will face,
  and use it to show that the code is able to recover the input power
  spectra in the most complex scenarios with no detectable
  bias. \nmt~can be found at \url{https://github.com/LSSTDESC/NaMaster}, and is provided with
  comprehensive documentation and a number of code examples to
  maximize its usefulness in the community.
\end{abstract}

@@

  \section{Introduction}\label{sec:intro}
  Two-point functions have proven to be the most useful summary
  statistic for cosmology in terms of data compression, due to the
  near-Gaussian nature of the modes of the cosmic density field that
  can be easily analyzed to extract cosmological
  information~\citep{1995PhRvL..74.4369B,1997PhRvD..55.5895T,2001PhRvD..64f3001T}. Since
  the only true observables in any astronomical dataset are the
  intensity and polarization as a function of frequency $\nu$ and sky
  position $\nv$, angular 2-point correlators play a particularly
  central role in most cosmological analyses, both in the form of
  configuration-space two-point correlation functions $\xi(\theta)$
  and as Fourier-space\footnote{In this paper we use the term
    ``Fourier'' to describe both the flat space Fourier expansion as
    well as harmonic-space expansion on the sphere. We will also use
    the term ``power spectrum'' to describe both the auto power
    spectrum of a field and cross-power spectra between two fields. }
  power spectra $C_\ell$
  \citep{2012MNRAS.427.1891A,2011PhRvD..84f3505B}.  Although
  correlation functions are often easier to estimate from complex
  observations~\citep{1993ApJ...412...64L}, a Fourier space analysis
  is advantageous for many other reasons. The most important is that
  Fourier space is a natural space for working with statistics of
  translation-invariant fields. Therefore, Fourier space analysis has
  a clearer separation between linear and non-linear scales, as well
  as Gaussian and non-Gaussian
  modes~\citep{2008PhRvD..77j3013H}. Moreover, individual band-power
  measurements are typically significantly less correlated compared to
  configuration-space analysis which makes covariance matrix
  estimation easier. This has motivated a large body of literature
  aimed at designing optimal and efficient power spectrum estimators.

    While maximum-likelihood~\citep{1998PhRvD..57.2117B,2003PhRvD..67b3001W} or minimum-variance and quadratic estimators~\citep{1997PhRvD..55.5895T,2000MNRAS.317L..23H,2001PhRvD..64f3001T} can recover the power spectrum with virtually no loss of information, their computational implementation is prohibitively expensive for high resolution data, since these methods scale as $\ell_{\rm max}^6$ and $\ell_{\rm max}^4$ respectively with the maximum multipole $\ell_{\rm max}$. For this reason, pseudo-$C_\ell$ algorithms \citep{1973ApJ...185..413P,2001PhRvD..64h3003W,2001ApJ...561L..11S,2001PhRvD..64h3003W,2002ApJ...567....2H,2002MNRAS.336.1304H,2004MNRAS.350..914C} have become a popular alternative that preserves a the $\ell_{\rm max}^3$ scaling of spherical harmonic transforms (and faster in the flat-sky approximation). This method can be used on fields with arbitrary spin \citep{2003ApJS..148..161K} in both curved and flat skies, has been extended to deal with contaminant deprojection in scalar fields \citep{2017MNRAS.465.1847E}, and can be optimized to measure the CMB $B$-mode power spectrum \citep{2002PhRvD..65b3505L,2003PhRvD..67b3501B,2006PhRvD..74h3002S,2009PhRvD..79l3515G,2011PhRvD..83h3003B}. Unfortunately, in spite of the advantages of this method, no public, validated and well-maintained codes exist supporting all of these capabilities. This paper presents a public software package, \nmt, that unifies and extends all of these features providing a general framework to estimate pseudo-$C_\ell$ angular power spectra in a wide variety of scenarios.

@@

        The pseudo-$C_\ell$ estimates will be optimal when these steps are close to what an optimal quadratic estimator would do. First, this requires that $\mC$ be close to diagonal, which is true when either the noise is large, or when the underlying power spectrum is close to white. Multiplication by the mask mixes modes, and replacement of multiplication by the response matrix with band-power averaging, only makes sense, when the mask is ``well-behaved'', i.e. compact with no high-frequency structure in it. If these conditions are not met, the pseudo-$C_\ell$ estimator will still be unbiased (by construction), but it will get progressively less optimal. In practical surveys, however, the loss of optimality rarely exceeds 10-20\% and is typically smaller~\citep{2013MNRAS.435.1857L}.

@@

      At this stage, the pseudo-$C_\ell$ method proceeds as usual, applying Eq. \ref{eq:pcldef} to the purified field. The purification process however requires a slight modification of the analytical form of the mode-coupling matrix. Specifically, if either $E$ or $B$ modes of a spin-2 field have been purified Eqs. \ref{eq:m02} and \ref{eq:m22} must be modified by exchanging one factor of
      \begin{equation}\nonumber
        \wtj{\ell}{\ell'}{\ell''}{2}{-2}{0}
      \end{equation}
      for
      \begin{align}\nonumber
        &\wtj{\ell}{\ell'}{\ell''}{2}{-2}{0}+2\frac{\alpha_{\ell,2}}{\alpha_{\ell,1}\alpha_{\ell'',1}}\wtj{\ell}{\ell'}{\ell''}{1}{-2}{1}\\
        &\,\,\,+\frac{\alpha_{\ell,2}}{\alpha_{\ell'',2}}\wtj{\ell}{\ell'}{\ell''}{0}{-2}{2}.
      \end{align}
      \emph{per purified field}. \todo{check this}.

@@

        \subsubsection{Beam deconvolution}\label{sssec:maths.beam_noise.beam}
          Including the effect of an spherically symmetric instrumental beam amounts to a simple redefinition of the mode-coupling matrix
          \begin{equation}
            {\sf M}^{s_as_b}_{\ell_1\ell_2}\rightarrow{\sf M}^{s_as_b}_{\ell_1\ell_2}W_{\ell_2}^{ab},
          \end{equation}
          where $W^{ab}_\ell$ is the product of the harmonic transform
          of the beams for maps $a$ and $b$. Note that due to hairy
          ball theorem, it is impossible to have an anisotropic beam
          that is uniform over the entire sphere.  However, it is
          possible that the beam is both anisotropic and spatially
          varying. In this case, the pseudo-$C_\ell$ power spetrum
          estimation will produce biased estimates that will be
          suppressed by an \emph{effective} beam. Correcting for the
          beam in such cases can also be done analytically, but we
          leave it for future work.  In practice, if required, the
          suppression factor can be found by applying the PCL
          estimator to a a set of mock input maps with the same
          underlying signal but  with and without application of the
          beam smoothing (to both data maps and contaminant maps) and taking
          the ration. \as{How does this interact with E/B purification?}

@@

    \subsection{Flat-sky pseudo-\texorpdfstring{$C_\ell$}{Cl}s}\label{ssec:maths.flat}
      Using the full-sky expression presented above and in Appendix \ref{app:shts} for the analysis of small sky patches, where the curvature of the sphere can be neglected, is numerically inefficient for two main reasons:
      \begin{enumerate}
        \item Full-sky data formats usually store information about the full sphere, even if the data being stored is limited to a small sky patch. This can lead to an inefficient use of memory and storage space, because the standard numerical implementations of spherical transforms operate on full-sky data.
        \item Spherical harmonic transforms are notoriously slower than discrete Fourier transforms (DFTs), both due to the scaling of both algorithms (${\cal O}(N_{\rm pix}^{3/2})$ for SHTs vs. ${\cal O}(N_{\rm pix}\log(N_{\rm pix}))$ for DFTs) and because by SHTs involve operations over the full sphere.
        \item Fourier transforms over a flat patch offer exact quadrature, which makes them inherently more numerically stable.
      \end{enumerate}
      For this reason, \nmt\, also supports the computation of power spectra from flat-sky maps. Appendix \ref{app:flat} describes the dictionary between spin-$s$ fields and their Fourier/harmonic coefficients defined in flat and curved skies, and we present the flat-sky versions of the pseudo-$C_\ell$ methods described in the previous sections here.

@@

f2=NmtField(mask,[map_q,map_u],
            templates=[[dust_q,dust_u]],
            purify_b=True,beam=beam_ell)
f0=NmtField(mask,[map_t])

b=NmtBin(nside,nlb=10)

wsp=NmtWorkspace()
wsp.compute_coupling_matrix(f0,f2,b)

cl_bias=deprojection_bias(f0,f2,cl_theory)

cl_coupled=compute_coupled_cell(f0,f2)

cl_decoupled=wsp.decouple_cell(cl_coupled,
                               cl_bias)
    \end{lstlisting}
    Briefly:
    \begin{enumerate}
      \item In line 1 we import \nmt's python module.
      \item All preliminary I/O operations (reading maps, masks etc.) take place implicitly in line 3.
      \item In line 5 we generate a field with spin 2. The field is defined in terms of a mask and maps of the $(Q,U)$ Stokes parameters. We also provide a set of contaminants to deproject, an instrumental beam and request for the field's $B$-modes to be purified. Deprojection and purification will take place at this stage in that order, as explained in Section \ref{ssec:maths.pureb}. The field's SHT is also estimated upon initialization.
      \item We define a second field in line 8. This one is a simpler spin-0 field with no contaminants or beam.
      \item We define the output bandpower structure in line 10. In this case we will use bins of 10 multipoles, although more general schemes are supported.
      \item In lines 12 and 13 we use an {\tt NmtWorkspace} object to compute the mode-coupling matrix.
      \item The deprojection bias is computed in line 15 using a best-guess for the true underlying power spectrum.
      \item The pseudo-$C_\ell$ is computed in line 17.
      \item The deconvolved bandpowers are finally estimated using the workspace in line 19.
    \end{enumerate}

>>>>

% Add your own macros here:
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{dsfont}
\usepackage{courier}

\lstset{ %
language=python,                % choose the language of the code
basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
numbers=left,                   % where to put the line-numbers
numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
frame=false,           % adds a frame around the code
tabsize=2,          % sets default tabsize to 2 spaces
captionpos=b,           % sets the caption-position to bottom
breaklines=true,        % sets automatic line breaking
breakatwhitespace=false,    % sets if automatic breaks should only happen at whitespace
escapeinside={\%*}{*)}          % if you want to add a comment within your code
}
%\usepackage[colorlinks,linkcolor=red,citecolor=blue,urlcolor=blue ]{hyperref}
 %\hypersetup{draft}
\newcommand{\nv}{\hat{\bf n}}
\newcommand{\wtj}[6]{\left(\begin{array}{ccc} #1 & #2 & #3\\#4 & #5 & #6\end{array} \right)}
\newcommand{\nmt}{{\tt NaMaster}}
\newcommand{\va}{{\mathbf{a}}}
\newcommand{\vb}{{\mathbf{b}}}
\newcommand{\mC}{{\sf C}}
\newcommand{\mM}{{\sf M}}
\newcommand{\mI}{{\mathds{1}}}

@@

\begin{abstract}
  The pseudo-$C_\ell$ is an algorithm for estimation of the angular
  power and cross-power spectra that is very fast and in realistic
  cases also nearly optimal. The algorithm can be extended to deal
  with contaminant deprojection and $E/B$ purification, and can
  therefore be applied in a wide variety of scenarios of interest for
  current and future cosmological observations. This paper presents
  \nmt, a public, validated, accurate and easy-to-use software
  package that, for the first time, provides a unified framework to
  compute angular cross-power spectra of any pair of spin-0 or spin-2
  fields, contaminated by an arbitrary number of linear systematics
  and requiring $B$- or $E$-mode purification, both in the sphere or
  in the flat-sky approximation. We describe the mathematical
  background of the estimator, including all the features above, and
  its software implementation in \nmt. We construct a validation suite
  that aims to resemble the types of observations that next-generation
  large-scale structure and ground-based CMB experiments will face,
  and use it to show that the code is able to recover the input power
  spectra in the most complex scenarios with no detectable
  bias. \nmt~can be found at \url{https://github.com/LSSTDESC/NaMaster}, and is provided with
  comprehensive documentation and a number of code examples.
\end{abstract}

@@

  \section{Introduction}\label{sec:intro}
  Two-point functions have proven to be the most useful summary
  statistic for cosmology in terms of data compression, due to the
  near-Gaussian nature of the modes of the cosmic density field that
  can be easily analyzed to extract cosmological
  information~\citep{1995PhRvL..74.4369B,1997PhRvD..55.5895T,2001PhRvD..64f3001T}. Since
  the only true observables in any astronomical dataset are the
  intensity and polarization as a function of frequency $\nu$ and sky
  position $\nv$, angular 2-point correlators play a particularly
  central role in most cosmological analyses, both in the form of
  configuration-space two-point correlation functions $\xi(\theta)$
  and as Fourier-space\footnote{In this paper we use the term
    ``Fourier'' to describe both the flat space Fourier expansion as
    well as harmonic-space expansion on the sphere. We will also use
    the term ``power spectrum'' to describe both the auto power
    spectrum of a field and cross-power spectra between two fields. }
  power spectra $C_\ell$
  \citep{2012MNRAS.427.1891A,2011PhRvD..84f3505B}.  Although
  correlation functions are often easier to estimate from complex
  observations~\citep{1993ApJ...412...64L}, a Fourier space analysis
  is advantageous for many other reasons. The most important is that
  Fourier space is a natural space for working with statistics of
  translation-invariant fields. Therefore, Fourier space analysis has
  a clearer separation between linear and non-linear scales, as well
  as Gaussian and non-Gaussian
  modes~\citep{2008PhRvD..77j3013H}. Moreover, individual band-power
  measurements are typically significantly less correlated compared to
  configuration-space analysis which makes covariance matrix
  estimation easier. This has motivated a large body of literature
  aimed at designing optimal and efficient power spectrum estimators and 
  used extensively in Cosmic Microwave Background (CMB) two-point measurements.

    While maximum-likelihood~\citep{1998PhRvD..57.2117B,2003PhRvD..67b3001W} or minimum-variance and quadratic estimators~\citep{1997PhRvD..55.5895T,2000MNRAS.317L..23H,2001PhRvD..64f3001T} can recover the power spectrum with virtually no loss of information, their computational implementation is prohibitively expensive for high resolution data, since these methods scale as $\ell_{\rm max}^6$ and $\ell_{\rm max}^4$ respectively with the maximum multipole $\ell_{\rm max}$. For this reason, pseudo-$C_\ell$ algorithms \citep{1973ApJ...185..413P,2001PhRvD..64h3003W,2001ApJ...561L..11S,2001PhRvD..64h3003W,2002ApJ...567....2H,2002MNRAS.336.1304H,2004MNRAS.350..914C} have become a popular alternative that preserves the $\ell_{\rm max}^3$ scaling of spherical harmonic transforms (and faster in the flat-sky approximation). This method can be used on fields with arbitrary spin \citep{2003ApJS..148..161K} in both curved and flat skies, has been extended to deal with contaminant deprojection in scalar fields \citep{2017MNRAS.465.1847E}, and can be optimized to measure the CMB $E/B$-mode power spectrum \citep{2002PhRvD..65b3505L,2003PhRvD..67b3501B,2006PhRvD..74h3002S,2009PhRvD..79l3515G,2011PhRvD..83h3003B}. Unfortunately, in spite of the advantages of this method, no public, validated and well-maintained codes exist supporting all of these capabilities. This paper presents a public software package, \nmt, that unifies and extends all of these features providing a general framework to estimate pseudo-$C_\ell$ angular power spectra in a wide variety of scenarios.

@@

        The pseudo-$C_\ell$ estimates will be optimal when these steps are close to what an optimal quadratic estimator would do. First, this requires that $\mC$ be close to diagonal, which is true when either the noise is large and uncorrelated, or when the underlying power spectrum is close to white. Multiplication by the mask mixes modes, and replacement of multiplication by the response matrix with band-power averaging, only makes sense, when the mask is ``well-behaved'', i.e. compact with no high-frequency structure in it. If these conditions are not met, the pseudo-$C_\ell$ estimator will still be unbiased (by construction), but it will get progressively less optimal. In practical surveys, however, the loss of optimality rarely exceeds 10-20\% and is typically smaller~\citep{2013MNRAS.435.1857L}.

@@

      At this stage, the pseudo-$C_\ell$ method proceeds as usual, applying Eq. \ref{eq:pcldef} to the purified field. The purification process however requires a slight modification of the analytical form of the mode-coupling matrix. Specifically, if either $E$ or $B$ modes of a spin-2 field have been purified Eqs. \ref{eq:m02} and \ref{eq:m22} must be modified by exchanging one factor of
      \begin{equation}\nonumber
        \wtj{\ell}{\ell'}{\ell''}{2}{-2}{0}
      \end{equation}
      for
      \begin{align}\nonumber
        &\wtj{\ell}{\ell'}{\ell''}{2}{-2}{0}+2\frac{\alpha_{\ell,2}}{\alpha_{\ell,1}\alpha_{\ell'',1}}\wtj{\ell}{\ell'}{\ell''}{1}{-2}{1}\\
        &\,\,\,+\frac{\alpha_{\ell,2}}{\alpha_{\ell'',2}}\wtj{\ell}{\ell'}{\ell''}{0}{-2}{2}.
      \end{align}
      per purified field. \todo{check this}.

@@

        \subsubsection{Beam deconvolution}\label{sssec:maths.beam_noise.beam}
          Including the effect of an spherically symmetric instrumental beam amounts to a simple redefinition of the mode-coupling matrix
          \begin{equation}
            {\sf M}^{s_as_b}_{\ell_1\ell_2}\rightarrow{\sf M}^{s_as_b}_{\ell_1\ell_2}W_{\ell_2}^{ab},
          \end{equation}
          where $W^{ab}_\ell$ is the product of the harmonic transform
          of the beams for maps $a$ and $b$. Note that since
          any continuous tangent vector field on the sphere 
          must, at some location, be zero (hairy ball theorem)
          it is impossible to have an anisotropic beam
          that is uniform over the entire sphere.  However, it is
          possible that the beam is both anisotropic and spatially
          varying. In this case, the pseudo-$C_\ell$ power spetrum
          estimation will produce biased estimates that will be
          suppressed by an \emph{effective} beam. Correcting for the
          beam in such cases can also be done analytically, but we
          leave it for future work.  In practice, if required, the
          suppression factor can be found by applying the PCL
          estimator to a a set of mock input maps with the same
          underlying signal but  with and without application of the
          beam smoothing (to both data maps and contaminant maps) and taking
          the ratio. \as{How does this interact with E/B purification?}

@@

    \subsection{Flat-sky pseudo-\texorpdfstring{$C_\ell$}{Cl}s}\label{ssec:maths.flat}
      Using the full-sky expression presented above and in Appendix \ref{app:shts} for the analysis of small sky patches, where the curvature of the sphere can be neglected, is numerically inefficient for two main reasons:
      \begin{enumerate}
        \item Full-sky data formats usually store information about the full sphere, even if the data being stored is limited to a small sky patch. This can lead to an inefficient use of memory and storage space, because the standard numerical implementations of spherical transforms operate on full-sky data.
        \item Spherical harmonic transforms are notoriously slower than discrete Fourier transforms (DFTs), both due to the scaling of both algorithms (${\cal O}(N_{\rm pix}^{3/2})$ for SHTs vs. ${\cal O}(N_{\rm pix}\log(N_{\rm pix}))$ for DFTs) and because by SHTs involve operations over the full sphere.
        \item Fourier transforms over a flat patch offer exact quadrature, which makes them inherently more numerically stable.
      \end{enumerate}
      For this reason, \nmt, also supports the computation of power spectra from flat-sky maps. Appendix \ref{app:flat} describes the dictionary between spin-$s$ fields and their Fourier/harmonic coefficients defined in flat and curved skies, and we present the flat-sky versions of the pseudo-$C_\ell$ methods described in the previous sections here.

@@

f2 = NmtField(mask, [map_q,map_u],
            templates = [[dust_q,dust_u]],
            purify_b = True,beam=beam_ell)
f0 = NmtField(mask, [map_t])

b = NmtBin(nside, nlb = 10)

wsp = NmtWorkspace()
wsp.compute_coupling_matrix(f0, f2, b)

cl_bias = deprojection_bias(f0, f2, cl_theory)

cl_coupled = compute_coupled_cell(f0, f2)

cl_decoupled = wsp.decouple_cell(cl_coupled, cl_bias)
    \end{lstlisting}
    Briefly:
    \begin{enumerate}
      \item In line 1 we import \nmt's python module.
      \item All preliminary I/O operations (reading maps, masks etc.) take place implicitly in line 3.
      \item In line 5 we generate a field with spin 2. The field is defined in terms of a mask and maps of the $(Q,U)$ Stokes parameters. We also provide a set of contaminants to deproject, an instrumental beam and request for the field's $B$-modes to be purified. Deprojection and purification will take place at this stage in that order, as explained in Section \ref{ssec:maths.pureb}. The field's SHT is also estimated upon initialization.
      \item We define a second field in line 8. This one is a simpler spin-0 field with no contaminants or beam.
      \item We define the output bandpower structure in line 10. In this case we will use bins of 10 multipoles, although more general schemes are supported.
      \item In lines 12 and 13 we use an {\tt NmtWorkspace} object to compute the mode-coupling matrix.
      \item The deprojection bias is computed in line 15 using a best-guess for the true underlying power spectrum.
      \item The pseudo-$C_\ell$ is computed in line 17.
      \item The deconvolved bandpowers are finally estimated using the workspace in line 19.
    \end{enumerate}
